Learning Discrete-valued Bayesian Networks from Mixed Data

### 2015-11-19 BN_discretizer_paper.tex
fixed some notation in the algorithms. Still need to go through that in detail

\DeclareMathOperator{\Pa}{pa} % parents of a var in a BN
\newcommand{\cX}{\mathcal{X}} % continuous X
\newcommand{\cx}{\mathcal{x}} % instance of x
\newcommand{\cu}{\mathcal{u}} % unique sorted instances of x

@@

% \begin{table}[H]
%   \centering
%   \begin{tabular}{@{}ccl@{}}
%     \toprule
%     Current & Mykel & Description \\
%     \midrule
%     % $G$ & $G$ & BN graph structure \\
%     % $P$ & $P$ & discrete probability \\
%     % $p$ & $p$ & probability density \\
%     % $D$ & $D$ & available data \\
%     $|D|$ & $n$ & number of variables in the BN \\
%     $D_{\bm{Y}}$ & & subset of data associated with $\bm{Y}$ \\
%     $X$ & $X_{1:n}$ & variables in BN (discrete) \\
%     $x_{1:n}$ & $x_{1:n}$ & a specific instance of vars in BN \\
%     $x'_{1:n}$ &  & unique instances of var in BN \\
%         & $r_i$ & number of instantiations of $i$th var \\
%     $J_p$ & $q_i$ & number of instantiations of parents of $i$th var \\
%         &       & $ = 1$ if no parents \\
%     $\brock{\Pi_{X_i}}_j$ & $\pi_{ij}$ & $j$th instantiation of $i$th var \\
%         & $\theta_{ijk}$ & probability of $i$th var $= k$ given $j$th parental inst. \\
%     $\Theta$ & $\bm{\theta}$ & set of all (non-independent) probability parameters \\
%     $n_{i,j}^{(k)}$ & $m_{ijk}$ & observation counts \\
%     $n_{i,j,l}^{(k)}$ & $m_{ijk}$ & observation counts with $l$th instance of spouse\\
%     $\alpha_{ijk}$ & $\alpha_{ijk}$ & Dirichlet prior counts \\
%     $N_{ij}$ & $\alpha_{ij0}$ & sum over Dirichlet prior counts for $k$ \\
%         & $m_{ij0}$ & sum over observations counts for $k$ \\
%         & $\Gamma$ & the Gamma function \\
%     $N$ &          & is the number of samples in dataset \\
%     $N_i$ &          & counts for $i$th instance of a specific var \\
%     $\Pi_X$ & $\text{pa}_{x_i}$ & parents of a variable or instance \\
%     $n_C$ & & number of children of a variable \\
%     $H(~)$ & & a function \\
%     $I(~,~)$ & & mutual information \\
%     $X_i$ & & $i$th variable in BN, may be cont or disc \\
%     $X_i^*$ & & discrete version of cont var \\
%     $E$ & & set of edges in BN \\
%     $\text{Pa}\paren{X_i}$, $n_p$ & & number of parents of $X_i$ \\
%     $f\paren{X_i, \Pi_{X_i}}$ & $f(G)$ & score function for K2 \\
%     $M_C$ & & discretization policy for $C$ \\
%     $C$ & & a continuous variable \\
%     $t_i$ & & the $i$th bin edge for a specific var \\
%     $d_i$ & & allowable discretization positions (between unique vals) \\
%     $k$ & & number of intervals for $M_C$ \\
%     $C_{1:N}$ & & sorted samples for $C$ \\
%     $n_i$ & & $n_1 = s_1$ and $n_i = s_{i} - s_{i-1}$ \\
%     $s_i$ & & $\sum_{j=1}^k n_j$ and $s_k = N$, $s_0 = 0$ \\
%     $C_i$ & & $i$th child of a specific var \\
%     $\bm{S}_i$ & & set of spouses associated with $i$th child of specific var \\
%     $L$ & & largest cardinality among discrete variables in BN \\
%     $I_q$ & & an interval from $u \rightarrow v$ \\
%     $v'_\text{max}$ & & largest cardinality of variables that directly connect to $X$ \\
%     $S[]$ & &  optimal objective value of subproblems with instances from $1$ to $u$ \\
%     \bottomrule
%   \end{tabular}
% \end{table}

@@

The MODL method uses dynamic programming to find the optimal discretization policy, and has an $\bigo\paren{N^3 + r \cdot N^2}$ runtime, where $N$ is the of number of data instances and $r$ is the number of instances for the class variable.
\cite{Lustgarten_2011} suggest several formulations for the prior term $P(\text{Model})$.
Under a uniform prior probability over discretizations the runtime can be reduced to $\bigo\paren{r \cdot N^2}$ without sacrificing optimality.

@@

\begin{small}
\begin{equation}
\label{eq:BayesianScore}
  \log P(D\mid G) = \log P(G) + \sum_{i=1}^n \sum_{j=1}^{q_i} \log \paren{
    \frac{
      \Gamma\paren{\alpha_{ij}^{(0)}}
    }{
      \Gamma\paren{\alpha_{ij}^{(0)} + \beta_{ij}^{(0)}}
    }
  } + \sum_{k=1}^{r_i} \log \paren{
    \frac{
      \Gamma\paren{\alpha_{ij}^{(k)} + \beta_{ij}^{(k)}}
    }{
      \Gamma\paren{\alpha_{ij}^{(k)}}
    }
  }\text{.}
\end{equation}
\end{small}

@@

\noindent
The discretization policy discretizes $\cX$ into $k$ intervals.
Let the samples of $\cX$ in a given dataset $D$ be sorted in ascending order, ${\cx_{1:n} = \curly{\cx_1 \leq \cx_2 \leq \ldots \leq \cx_n}}$, and let the unique values be ${\cu_{1:n'} = \curly{\cu_1 \leq \cu_2 \leq \ldots \leq \cu_{n'}}}$.
The index of the last occurence of $\cu_i$ in $\cx_{1:n}$ is denoted $s_i$.

@@

\noindent
where $\lambda_1 = s_1$, $\lambda_i \in s_{1:n'}$, $\gamma_1 = \lambda_1$, and $\gamma_i = \lambda_i - \lambda_{i-1}$.
The $\lambda_{1:k}$ representation is the number of instances before every discretization edge whereas the $\lambda$ representation is the number of instances within each discretization interval.

@@

% Let $D$ be a dataset of $N$ samples from which the discretization will be learned, sorted in ascending order according to $X$.
% Let $D_{\boldsymbol{Y}}$ will refer to the data instances associated with variables $\boldsymbol{Y}$, thus $D_X = \{ x_1,x_2,x_3,\ldots,x_N \}$.
% In order to facilitate understanding, we temporarily assume all values in $D_X$ are unique, that is, $x_1 < x_2 < \ldots < x_N$.
% It follows that a discretization policy $\Lambda$ on $X$ can be written as $M_X = \brock{n_1,n_2,\ldots,n_k}$, where $k$, $n_1$, $n_2$, $\ldots,n_k$ are positive integers satisfying $N = \sum_{i=1}^k n_i$.

@@

From the first principle one obtains the $P(\Lambda)$ of $\Lambda$ described in Equation \ref{eq:disc_def}
\begin{equation}
  \label{eq:p_M}
  P(\Lambda) = \prod_{i=1}^{k-1}
    \brock{
      1 - \exp\paren{
        - L \cdot \frac{
                         \cx_{\lambda_i +1} - \cx_{\lambda_i}
                       }{
                         \cx_n - \cx_1
                        }
      }
    }     
    \prod_{i=1}^{k}
	\brock{    
    \exp\paren{
      -L \cdot \frac{\cx_{\lambda_{i}} - \cx_{\lambda_{i-1} + 1}}{\cx_n - \cx_1}}
    }\text{.}
\end{equation}

@@

\begin{equation}
\label{eq:opt_prob}
\begin{aligned}
  & \sum_{i=1}^{k-1}
   - \log
    \paren{
      1 - \exp
      \paren{
        - L \cdot  \frac{
                         \cx_{\lambda_i +1} - \cx_{\lambda_i}
                       }{
                         \cx_n - \cx_1
                        }
      }
    }
      + \sum_{i=1}^{k}
      L \cdot \frac{\cx_{\lambda_{i}} - \cx_{\lambda_{i-1} + 1}}{\cx_n - \cx_1} \\
  & \sum_{i=1}^{k} \brock{
  \log{{\gamma_i + J_P - 1}\choose{J_P - 1}}
  +{ \log \paren{
        \frac{
          {\gamma_i}!
        }{
          {n^{({P})}_{i,1}}! \; {n^{({P})}_{i,2}}! \; \cdots \; {n^{({P})}_{i,J_P}}!
        }
      }}
  } \\
  & \sum_{j=1}^{n_c} \sum_{i=1}^k \sum_{l=1}^{J^{(j)}_S} \brock{
  { \log
          {{n^{(j)}_{i,l} + J_{C_j} - 1}\choose{J_{C_j}-1}}}        
    +
    \log \paren{ {
          \frac{
            {n^{(j)}_{i,l}}!
          }{
            {n^{(j)}_{i,1,l}!} \; {n^{(j)}_{i,2,l}!} \; \cdots \; {n^{(j)}_{i,J_{C_j},l}!}
          }
        }        
        } }
\end{aligned}
\end{equation}

@@

\begin{small}
\begin{equation}
\begin{aligned}
h(u,v) &=  \log {{\gamma_{q} + J_P - 1}\choose{J_P-1}} + \log \left( { {{\gamma_q}!}\over{ {n^{(p)}_{q,1} !} {n^{(p)}_{q,2} !} \cdots {n^{(p)}_{q,J_p} !}} } \right) \\
& + \sum_{j=1}^{n_c} \sum_{l=1}^{J^{(j)}_S} \brock{
  { \log
          {{n^{(j)}_{i,l} + J_{C_j} - 1}\choose{J_{C_j}-1}}}        
    +
    \log \paren{ {
          \frac{
            {n^{(j)}_{i,l}}!
          }{
            {n^{(j)}_{i,1,l}!} \; {n^{(j)}_{i,2,l}!} \; \cdots \; {n^{(j)}_{i,J_{C_j},l}!}
          }
        }        
        } }
\end{aligned}
\end{equation}
\end{small}

The evaluation of $h(u,v)$ for all $u \leq v$ is summarized in Algorithm \ref{alg:h} in the appendix.
The calculation has a $\bigo \paren{ n_c  {L'} \cdot N^2 + {L'}^{n_p} \cdot N^2 }$ runtime, where $n_c$ and $n_p$ are the numbers of child and parent variables respectively, and $L'$ is the largest cardinality of variables that directly connect to $X$. \tim{so parents and children, or parents, children and spouses?}

@@

\begin{algorithm}
\caption{Discretization of one continuous variable in a Bayesian network}
\label{alg:disc_one}
\begin{algorithmic}[5]
\Function{DiscretizeOne}{$\cX$, $D$, $G$}
\State $n \leftarrow$ the number of data instances
\State $n' \leftarrow$ the number of unique values in $D_{\cX}$
\State $H \leftarrow$ an $N \times N$ matrix such that $H[u,v] = h(u,v)$ (see Algorithm~\ref{alg:h} in Appendix)
\State $s_{1:n'} \leftarrow$ an increasing sequence of integers defined in Section \ref{subsec:disc_policy}
\State $L \leftarrow$ the largest cardinality over all discrete variables in the Markov blanket of $X$
\State $S[u] \leftarrow$ the optimal objective value of subproblems from instance $1$ to $s_u$
\State $\Lambda[u] \leftarrow$ the optimal discretization for subproblems from instance $1$ to $s_u$
\State $W[u]  \leftarrow - \log\brock{1 - {\exp\paren{- L \cdot{ {{x_{s_u+1} - x_{s_u}}\over{x_N - x_1}}}}}}$ for $u = 1,2, \ldots,n'-1$ and $W[n] \leftarrow 0$
\For {$v \leftarrow 1$ to $n'$}
\If {$v = 1$}
\State $S[v] \leftarrow H \left(1,s_v \right) + L[v]$
\State $\Lambda[v] \leftarrow \{ ({x_{s_v} + x_{s_v+1}}) / 2\}$
\Else
\State $s \leftarrow \infty$
\State $\text{DiscEdge} \leftarrow \infty$
\For {$u \leftarrow 1$ to $v$}
\State $s' \leftarrow S[u] + H \left( s_u +1,s_v \right) +  {L \cdot {{x_{s_v} - x_{s_u + 1}}\over{x_N - x_1}}} + W[v]$
\If {$s' < s$}
\State $s \leftarrow s'$
\State $\text{DiscEdge} \leftarrow ({x_{s_u} + x_{s_u+1}}) / 2$
\EndIf
\EndFor
\State $S[v] \leftarrow s$
\State $\Lambda[v] \leftarrow M[u] \cup \{ DiscEdge \}$
\EndIf
\EndFor
 \State \Return $\Lambda[n']$
\EndFunction
\end{algorithmic}
\end{algorithm}

@@

\begin{small}
\begin{equation}
\label{eq:opt_prob_approx}
\begin{aligned}
  & \sum_{i=1}^{k-1}
   - \log
    \paren{
      1 - \exp
      \paren{
        - L \cdot  \frac{
                         \cx_{\lambda_i +1} - \cx_{\lambda_i}
                       }{
                         \cx_n - \cx_1
                        }
      }
    }
      + \sum_{i=1}^{k}
      L \cdot \frac{\cx_{\lambda_{i}} - \cx_{\lambda_{i-1} + 1}}{\cx_n - \cx_1} \\
  & \sum_{i=1}^{k} \brock{
  \log{{\gamma_i + J_P - 1}\choose{J_P - 1}}
  +{ \log \paren{
        \frac{
          {\gamma_i}!
        }{ \sum_{r=1}^{n_p}
          {n^{({P_r})}_{i,1}}! \; {n^{({P_r})}_{i,2}}! \; \cdots \; {n^{({P_r})}_{i,J_{P_r}}}!
        }
      }}
  } \\
  & \sum_{j=1}^{n_c} \sum_{i=1}^k \sum_{l=1}^{J^{(j)}_S} \brock{
  { \log
          {{n^{(j)}_{i,l} + J_{C_j} - 1}\choose{J_{C_j}-1}}}        
    +
    \log \paren{ {
          \frac{
            {n^{(j)}_{i,l}}!
          }{
            {n^{(j)}_{i,1,l}!} \; {n^{(j)}_{i,2,l}!} \; \cdots \; {n^{(j)}_{i,J_{C_j},l}!}
          }
        }        
        } }
\end{aligned}
\end{equation}
\end{small}

The approximation reduces the runtime of Algorithm~\ref{alg:disc_one} to a polynomial, $\bigo\paren{ {v'_\text{max}} \cdot (n_c + n_p) \cdot N^2}$.
It will be shown in Section~\ref{sec:experiments} that the approximation is more sensitive to the distribution over other variables and is biased towards slightly higher interval counts.
\tim{We need to have a more specific thing to say along the lines of When does it fail?}

@@

\begin{small}
  \begin{equation}
  \begin{aligned}
  \label{eq:example_approx}
   \sum_{i=1}^k & \left\lbrace   \left[ \log{{\gamma_{i} + J_{C_1} - 1}\choose{J_{C_1}-1}} + \log \left(  {{{\gamma_i}!} \over { {n^{(1)}_{i,1,\emptyset} !} \cdots {n^{(1)}_{i,J_{C_1},\emptyset} !}} }  \right)  +  \right. \right.\\
  & \left.  \log{{\gamma_{i} + J_{C_2} - 1}\choose{J_{C_2}-1}} + \log \left(  {{{n_i}!} \over { {n^{(2)}_{i,1,\emptyset} !} \cdots {n^{(2)}_{i,J_{C_1},\emptyset} !}} }  \right)  \right] + \\
  &  \left. \left[  \log {{\gamma_{i} + J_P - 1}\choose{J_P-1}} +  \log \left( { {{{\gamma}_i}!}\over{ {n^{(P_1)}_{i,1} !}\cdots {n^{(P_1)}_{i,J_{P_1}} !}}} \right) +\log \paren{{ {{{\gamma}_i}!}\over{ {n^{(P_2}_{i,1} !}\cdots {n^{(P_2)}_{i,J_{P_2}} !}}} } \right] \right\rbrace \text{.}
  \end{aligned}
  \end{equation}
\end{small}

In Equation~\ref{eq:example_approx}, each child carries two terms, as shown in the first set of square brackets, but the two parent variables only carry a total of three terms, as shown in the second set of square brackets.
In this case the child variables have greater effect than the parent variables even though the number of child and parent variables are the same.
However, if $C_2$ has an additional parent $S_2$, as shown in the right graph of Figure~\ref{fig:example_networks}, the importances of $C_2$ will be reduced, because the information from $C_2$ is now affected by $S_2$.
This argument shows that the proposed method, both before and after approximation, indeed incorporates graph structure when discretizing continuous variables.

@@

\section{Multiple Variable Discretization}
\label{sec:multi_var}

@@

The single-variable discretization method can be extended to Bayesian networks with multiple continuous variables by iteratively discretizing variables individually.
The discretization process for a single variable requires that all other variables be discrete.
The iterative approach thus requires a prediscretization policy from which to initialize the process.
This work uses prediscretization with equal-width intervals, defined by

\begin{equation}
M_X = \curly{X_\text{min}, X_\text{min} + \delta, X_\text{min} + 2\delta, \ldots, X_\text{max}}\text{,}
\end{equation}

\noindent
where $X_\text{min}$ and $X_\text{max}$ are the extrema values of $D_X$, $\delta = (X_\text{max} - X_\text{min})/k$, and $k$ is the desired number of intervals.
This work sets $k$ to the median value of the cardinalities of initially discrete variables in the network.
After prediscretization, the one-variable discretization methis is iteratively applied over each continuous variable in reverse topoligical order, from the leaves to the root.
Let one such series of discretizations be a cycle.
Reverse topological order has the advantage of using less prediscretized instantiations during the first cycle.
For example, in the right-hand graph of Figure \ref{fig:example_networks}, assume $S_2$ is the only discrete variable.
The discretization of $P_1$ involves both $P_2$ and $X$, whereas the discretization of $C_1$ only involves $X$.

The algorithm is terminated when the number of discretization intervals and their associated edges converge for all variables.
A maximum cycle count is enforced to prevent infinite iterations when convergence does not occur.
The algorithm typically converges within a few cycles when tested on real-world data.
In cases where it does not converge, ten cycles usually produce decent discretization results.

The pseudocode for the multi-variable discretization procedure is shown in Algorithm \ref{alg:disc_two}.
It requires four inputs: $D$, the data instances from all network variables; $G$, the network structure; $C$, the set of all continuous variables in a reverse topological order, and $u_{cycle}$, the upper bound of times of cycles.

>>>>

\DeclareMathOperator{\Pa}{pa} % parents of a var in a BN
\newcommand{\cX}{\mathcal{X}} % continuous X
\newcommand{\cx}{\mathcal{x}} % instance of x
\newcommand{\cu}{\mathcal{u}} % unique sorted instances of x
\newcommand{\maxpasses}{\hat{n}_\text{cycle}}

@@

@@

The MODL method uses dynamic programming to find the optimal discretization policy, and has an $\bigo\paren{n^3 + r \cdot n^2}$ runtime, where $n$ is the of number of data instances and $r$ is the number of instances for the class variable.
\cite{Lustgarten_2011} suggest several formulations for the prior term $P(\text{Model})$.
Under a uniform prior probability over discretizations the runtime can be reduced to $\bigo\paren{r \cdot n^2}$ without sacrificing optimality.

@@

\begin{small}
\begin{equation}
\label{eq:BayesianScore}
  \ln P(D\mid G) = \ln P(G) + \sum_{i=1}^n \sum_{j=1}^{q_i} \ln \paren{
    \frac{
      \Gamma\paren{\alpha_{ij}^{(0)}}
    }{
      \Gamma\paren{\alpha_{ij}^{(0)} + \beta_{ij}^{(0)}}
    }
  } + \sum_{k=1}^{r_i} \ln \paren{
    \frac{
      \Gamma\paren{\alpha_{ij}^{(k)} + \beta_{ij}^{(k)}}
    }{
      \Gamma\paren{\alpha_{ij}^{(k)}}
    }
  }\text{.}
\end{equation}
\end{small}

@@

\noindent
The discretization policy discretizes $\cX$ into $k$ intervals.
Let the samples of $\cX$ in a given dataset $D$ be sorted in ascending order, ${\cx_{1:n} = \curly{\cx_1 \leq \cx_2 \leq \ldots \leq \cx_n}}$, and let the unique values be ${\cu_{1:m} = \curly{\cu_1 \leq \cu_2 \leq \ldots \leq \cu_{m}}}$.
The index of the last occurence of $\cu_i$ in $\cx_{1:n}$ is denoted $s_i$.

@@

\noindent
where $\lambda_1 = s_1$, $\lambda_i \in s_{1:m}$, $\gamma_1 = \lambda_1$, and $\gamma_i = \lambda_i - \lambda_{i-1}$.
The $\lambda_{1:k}$ representation is the number of instances before every discretization edge whereas the $\lambda$ representation is the number of instances within each discretization interval.

@@

% Let $D$ be a dataset of $N$ samples from which the discretization will be learned, sorted in ascending order according to $X$.
% Let $D_{\boldsymbol{Y}}$ will refer to the data instances associated with variables $\boldsymbol{Y}$, thus $D_X = \{ x_1,x_2,x_3,\ldots,x_n \}$.
% In order to facilitate understanding, we temporarily assume all values in $D_X$ are unique, that is, $x_1 < x_2 < \ldots < x_n$.
% It follows that a discretization policy $\Lambda$ on $X$ can be written as $M_X = \brock{n_1,n_2,\ldots,n_k}$, where $k$, $n_1$, $n_2$, $\ldots,n_k$ are positive integers satisfying $N = \sum_{i=1}^k n_i$.

@@

From the first principle one obtains the $P(\Lambda)$ of $\Lambda$ described in Equation \ref{eq:disc_def}

\begin{small}
\begin{equation}
  \label{eq:p_M}
  P(\Lambda) = \prod_{i=1}^{k-1}
    \brock{
      1 - \exp\paren{
        - L \cdot \frac{
                         \cx_{\lambda_i +1} - \cx_{\lambda_i}
                       }{
                         \cx_n - \cx_1
                        }
      }
    }
    \prod_{i=1}^{k}
	\brock{
    \exp\paren{
      -L \cdot \frac{\cx_{\lambda_{i}} - \cx_{\lambda_{i-1} + 1}}{\cx_n - \cx_1}}
    }\text{.}
\end{equation}
\end{small}

@@

\begin{equation}
\label{eq:opt_prob}
\begin{aligned}
  & \sum_{i=1}^{k-1}
   - \ln
    \paren{
      1 - \exp
      \paren{
        - L \cdot  \frac{
                         \cx_{\lambda_i +1} - \cx_{\lambda_i}
                       }{
                         \cx_n - \cx_1
                        }
      }
    }
      + \sum_{i=1}^{k}
      L \cdot \frac{\cx_{\lambda_{i}} - \cx_{\lambda_{i-1} + 1}}{\cx_n - \cx_1} \\
  & \sum_{i=1}^{k} \brock{
  \ln{{\gamma_i + J_P - 1}\choose{J_P - 1}}
  +{ \ln \paren{
        \frac{
          {\gamma_i}!
        }{
          {n^{({P})}_{i,1}}! \; {n^{({P})}_{i,2}}! \; \cdots \; {n^{({P})}_{i,J_P}}!
        }
      }}
  } \\
  & \sum_{j=1}^{n_c} \sum_{i=1}^k \sum_{l=1}^{J^{(j)}_S} \brock{
  { \ln
          {{n^{(j)}_{i,l} + J_{C_j} - 1}\choose{J_{C_j}-1}}}
    +
    \ln \paren{ {
          \frac{
            {n^{(j)}_{i,l}}!
          }{
            {n^{(j)}_{i,1,l}!} \; {n^{(j)}_{i,2,l}!} \; \cdots \; {n^{(j)}_{i,J_{C_j},l}!}
          }
        }
        } }
\end{aligned}
\end{equation}

@@

\begin{small}
  \begin{equation}
  \begin{aligned}
  h(u,v) &=  \ln {{\gamma_{q} + J_P - 1}\choose{J_P-1}} + \ln \left( { {{\gamma_q}!}\over{ {n^{(p)}_{q,1} !} {n^{(p)}_{q,2} !} \cdots {n^{(p)}_{q,J_p} !}} } \right) \\
  & + \sum_{j=1}^{n_c} \sum_{l=1}^{J^{(j)}_S} \brock{
    { \ln
            {{n^{(j)}_{i,l} + J_{C_j} - 1}\choose{J_{C_j}-1}}}
      +
      \ln \paren{ {
            \frac{
              {n^{(j)}_{i,l}}!
            }{
              {n^{(j)}_{i,1,l}!} \; {n^{(j)}_{i,2,l}!} \; \cdots \; {n^{(j)}_{i,J_{C_j},l}!}
            }
          }
          } }
  \end{aligned}
  \end{equation}
\end{small}

The evaluation of $h(u,v)$ for all $u \leq v$ is summarized in Algorithm~\ref{alg:h} in the appendix.
The calculation has a $\bigo \paren{ n_c  {L'} \cdot N^2 + {L'}^{n_p} \cdot N^2 }$ runtime, where $n_c$ and $n_p$ are the numbers of child and parent variables respectively, and $L'$ is the largest cardinality of variables that directly connect to $X$. \tim{so parents and children, or parents, children and spouses?}

@@

\begin{algorithm}
  \caption{Discretization of one continuous variable in a Bayesian network}
  \label{alg:disc_one}
  \begin{algorithmic}[5]
    \Function{DiscretizeOne}{$\cX$, $D$, $G$}
      \State $n \leftarrow$ the number of data instances
      \State $m \leftarrow$ the number of unique values in $D_{\cX}$
      \State $H \leftarrow$ an $N \times N$ matrix such that $H[u,v] = h(u,v)$ (see Algorithm~\ref{alg:h} in Appendix)
      \State $s_{1:m} \leftarrow$ an increasing sequence of integers defined in Section \ref{subsec:disc_policy}
      \State $L \leftarrow$ the largest cardinality over all discrete variables in the Markov blanket of $X$
      \State $S[u] \leftarrow$ the optimal objective value of subproblems from instance $1$ to $s_u$
      \State $\Lambda[u] \leftarrow$ the optimal discretization for subproblems from instance $1$ to $s_u$
      \State $W[u]  \leftarrow - \ln\brock{1 - {\exp\paren{- L \cdot{ {{\cx_{s_u+1} - \cx_{s_u}}\over{\cx_n - \cx_1}}}}}}$ for $u = 1,2, \ldots,m-1$ and $W[n] \leftarrow 0$
      \For {$v \leftarrow 1$ to $m$}
        \If {$v = 1$}
          \State $S[v] \leftarrow H \paren{1,s_v} + L[v]$
          \State $\Lambda[v] \leftarrow \{ ({\cx_{s_v} + \cx_{s_v+1}}) / 2\}$
        \Else
          \State $s \leftarrow \infty$
          \State $\text{DiscEdge} \leftarrow \infty$
        \For {$u \leftarrow 1$ to $v$}
        \State $s' \leftarrow S[u] + H \paren{s_u+1, s_v} +  {L \cdot {\frac{\cx_{s_v} - \cx_{s_u + 1}}{\cx_n - \cx_1}}} + W[v]$
        \If {$s' < s$}
          \State $s \leftarrow s'$
          \State $\text{DiscEdge} \leftarrow ({x_{s_u} + x_{s_u+1}}) / 2$
        \EndIf
        \EndFor
          \State $S[v] \leftarrow s$
          \State $\Lambda[v] \leftarrow M[u] \cup \{ DiscEdge \}$
        \EndIf
      \EndFor
      \State \Return $\Lambda[m]$
    \EndFunction
  \end{algorithmic}
\end{algorithm}

@@

\begin{small}
  \begin{equation}
  \label{eq:opt_prob_approx}
  \begin{aligned}
    & \sum_{i=1}^{k-1}
     - \ln
      \paren{
        1 - \exp
        \paren{
          - L \cdot  \frac{
                           \cx_{\lambda_i +1} - \cx_{\lambda_i}
                         }{
                           \cx_n - \cx_1
                          }
        }
      }
        + \sum_{i=1}^{k}
        L \cdot \frac{\cx_{\lambda_{i}} - \cx_{\lambda_{i-1} + 1}}{\cx_n - \cx_1} \\
    & \sum_{i=1}^{k} \brock{
    \ln{{\gamma_i + J_P - 1}\choose{J_P - 1}}
    +{ \ln \paren{
          \frac{
            {\gamma_i}!
          }{ \sum_{r=1}^{n_p}
            {n^{({P_r})}_{i,1}}! \; {n^{({P_r})}_{i,2}}! \; \cdots \; {n^{({P_r})}_{i,J_{P_r}}}!
          }
        }}
    } \\
    & \sum_{j=1}^{n_c} \sum_{i=1}^k \sum_{l=1}^{J^{(j)}_S} \brock{
    { \ln
            {{n^{(j)}_{i,l} + J_{C_j} - 1}\choose{J_{C_j}-1}}}
      +
      \ln \paren{ {
            \frac{
              {n^{(j)}_{i,l}}!
            }{
              {n^{(j)}_{i,1,l}!} \; {n^{(j)}_{i,2,l}!} \; \cdots \; {n^{(j)}_{i,J_{C_j},l}!}
            }
          }
          } }
  \end{aligned}
  \end{equation}
\end{small}

The approximation reduces the runtime of Algorithm~\ref{alg:disc_one} to a polynomial, $\todo{\bigo\paren{ {v'_\text{max}} \cdot (n_c + n_p) \cdot n^2}}$.
It will be shown in Section~\ref{sec:experiments} that the approximation is more sensitive to the distribution over other variables and is biased towards slightly higher interval counts.
\tim{We need to have a more specific thing to say along the lines of `When does it fail?'}

@@

\begin{small}
  \begin{equation}
  \begin{aligned}
  \label{eq:example_approx}
   \sum_{i=1}^k & \left\lbrace   \left[ \ln{{\gamma_{i} + J_{C_1} - 1}\choose{J_{C_1}-1}} + \ln \left(  {{{\gamma_i}!} \over { {n^{(1)}_{i,1,\emptyset} !} \cdots {n^{(1)}_{i,J_{C_1},\emptyset} !}} }  \right)  +  \right. \right.\\
  & \left.  \ln{{\gamma_{i} + J_{C_2} - 1}\choose{J_{C_2}-1}} + \ln \left(  {{{n_i}!} \over { {n^{(2)}_{i,1,\emptyset} !} \cdots {n^{(2)}_{i,J_{C_1},\emptyset} !}} }  \right)  \right] + \\
  &  \left. \left[  \ln {{\gamma_{i} + J_P - 1}\choose{J_P-1}} +  \ln \left( { {{{\gamma}_i}!}\over{ {n^{(P_1)}_{i,1} !}\cdots {n^{(P_1)}_{i,J_{P_1}} !}}} \right) +\ln \paren{{ {{{\gamma}_i}!}\over{ {n^{(P_2}_{i,1} !}\cdots {n^{(P_2)}_{i,J_{P_2}} !}}} } \right] \right\rbrace \text{.}
  \end{aligned}
  \end{equation}
\end{small}

In Equation~\ref{eq:example_approx}, each child carries two terms, as shown in the first set of square brackets, but the two parent variables only carry a total of three terms, as shown in the second set of square brackets.
In this case the child variables have greater effect on the optimal discretization than the parent variables, even though the number of child and parent variables are the same.
However, if $C_2$ has an additional parent $S_2$, as shown in the right graph of Figure~\ref{fig:example_networks}, the importances of $C_2$ will be reduced.
This is due to the additional influence of $S_2$ on $C_2$.
This phenomenon arrises because the Bayesian method, but with and without the heuristic, incorporates graph structure when discretizing continuous variables.

@@

\section{Multi-Variable Discretization}
\label{sec:multi_var}

@@

The single-variable discretization method can be extended to Bayesian networks with multiple continuous variables by iteratively discretizing individual variables.
The discretization process for a single variable requires that all other variables be discrete.
The iterative approach uses a prediscretization policy in order to initialize the process.
This work prediscretizes all continuous variables into $k$ equal-width intervals, where $k$ is the median value of the number of intervals of initially discrete variables in the network.

After prediscretization, the one-variable discretization method is iteratively applied over each continuous variable in reverse topoligical order, from the leaves to the root.
Reverse topological order has the advantage of relying on fewer prediscretizations of the continuous variables during the first pass.
For example, in the right-hand graph of Figure~\ref{fig:example_networks}, assume $S_2$ is the only discrete variable.
The discretization of $P_1$ involves both $P_2$ and $\cX$, whereas the discretization of $C_1$ only involves $\cX$.

The algorithm is terminated when the number of discretization intervals and their associated edges converge for all variables, and a maximum number of complete passes is enforced to prevent infinite iterations when convergence does not occur.
The algorithm typically converges within a few passes when tested on real-world data.
In cases where it does not converge, \tim{when was this observed? What dataset?}, ten passes usually produce decent discretization results.

The pseudocode for the multi-variable discretization procedure is shown in Algorithm~\ref{alg:disc_two}.
It requires four inputs: $D$, a dataset of samples from the joint distribution; $G$, the fixed network structure; $\bm{\cX}$, the set of all continuous variables in reverse topological order, and $\maxpasses$, an upper bound on the number of complete passes.

\begin{algorithm}
  \caption{Discretization of multiple continuous variables \todo{Fix label reference number}}
  \label{alg:disc_two}
  \begin{algorithmic}[5]
  \Function{DiscretizeAll}{$D$, $G$, $\bm{\cX}$, $\maxpasses$}
    \State $\Lambda_{\bm{\cX}} \leftarrow$ the discretization policies for each $\cX$ in $\bm{\cX}$
    \State $D^* \leftarrow $ the dataset $D$ discretized according to $\Lambda_{\bm{\cX}}$
    \State $k \leftarrow \textsc{Median}\paren{\curly{ |X| \text{ s.t. } X \text{ does not have corresponding } \cX \text{ in } \bm{\cX} }}$
    \For {$\paren{X, \cX}$ such that $\cX \in \bm{\cX}$}
      \State $\Lambda_{\cX} \leftarrow$ equal-width discretization of $\cX$ with $k$ intervals
      \State $D^*_{X} \leftarrow \Lambda_{\cX} \paren{D_{\cX}}$
    \EndFor
    \State $n_\text{cycle} \leftarrow 0$
    \While {$\Lambda_{\bm{\cX}}$ has not converged \textbf{and} $n_\text{cycle} \leq \maxpasses$}
      \State increment {$n_\text{cycle}$}
      \For {$\paren{X, \cX}$ such that $\cX \in \bm{\cX}$}
        \State {$D^*_X \leftarrow D_{\cX}$}
        \State {$\Lambda_{\cX} \leftarrow \textsc{DiscretizeOne}(\cX, D^*, G)$}
        \State {$D^*_X \leftarrow \Lambda_{\cX}\paren{D_{\cX}}$}
      \EndFor
    \EndWhile
    \State \Return $\Lambda_{\bm{\cX}}$
  \EndFunction
  \end{algorithmic}
\end{algorithm}
